Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-14 _tex/paper2.tex
Built site for gh-pages

@@

@@

@@

@@

@@

@@

Researchers would then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that transforms data into an estimate \citep{Neal_2020}.
\emph{Estimates} are numerical values that approximate the estimand and
are derived through \emph{estimation}, which refers to the process of
integrating data with an estimator \citep{Everitt_et_al_2010}. The
Identification-Estimation flowchart \citep{McElreath_2020, Neal_2020} in
Figure~\ref{fig-IEflow} provides a visual representation of the process
of transitioning from estimands to estimates.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. In
essence, SCMs and DAGs act as \emph{conceptual models} that guide
researchers in determining which statistical models can yield valid
causal inferences, assuming the depicted causal structure of the models
is correct \citep{McElreath_2020}. Notably, every SCM has an associated
DAG \citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a
visual representation of the role of theoretical models in the inference
process.

SCMs and DAGs offer two key advantages for modeling causal structures.
First, they enable the representation of causal relations in a
non-parametric and fully interactive manner. This feature allows for
feasible ACE identification strategies without specifying the data type
or the nature of the functional dependence among variables
\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
represent a wide range of causal structures using only five fundamental
building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures, facilitating their analysis
by focusing on the causal assumptions associated with each building
block \citep{McElreath_2020}.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\)
\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
are those whose causal mechanisms a researcher chooses to model
\citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

The right panels of the figures display the complementary DAGs. A DAG
consists of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} means the edges extend from one node
to another, with arrows indicating the direction of causal influence.
The term \emph{acyclic} signifies that the causal influences do not form
loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
expected direction of causal influences among these variables. While
DAGs often omit exogenous variables for simplicity (the \emph{standard}
representation), including these variables is advantageous (the
\emph{magnified} representation shown in the figures), as their
inclusion helps to identify potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

@@

A careful examination of the figures highlights the assumptions
underlying these building blocks. Figures \ref{fig-scm_bb1} and
\ref{fig-mdag_bb1} depict two unconnected nodes, representing a scenario
where variables \(X_{1}\) and \(X_{3}\) are not causally related.
Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
connected nodes, representing a scenario where a \emph{parent} node
\(X_{1}\) exerts a causal influence on a \emph{child} node \(X_{3}\). In
this setup, \(X_{3}\) is considered a \emph{descendant} of \(X_{1}\).
Additionally, \(X_{1}\) and \(X_{3}\) are described as \emph{adjacent}
because an edge directly connects them. Figures \ref{fig-scm_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X_{1}\) influences
\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In this configuration,
\(X_{1}\) is a parent node of \(X_{2}\), which is a parent node of
\(X_{3}\). This structure creates a \emph{directed path} between
\(X_{1}\) and \(X_{3}\). Consequently, \(X_{1}\) is an \emph{ancestor}
of \(X_{3}\), and \(X_{2}\) fully \emph{mediates} the relationship
between the two. Figures \ref{fig-scm_bb4} and \ref{fig-mdag_bb4}
illustrate a \emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are
both influenced by \(X_{2}\). Here, \(X_{2}\) is a parent node that
\emph{confounds} the relationship between \(X_{1}\) and \(X_{3}\).
Finally, Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a
\emph{collider}, where variables \(X_{1}\) and \(X_{3}\) are concurrent
causes of \(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are
not causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, in all SCMs, the errors are assumed to be
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \perp\!\!\!\perp e_{X2}\),
\(e_{X1} \perp\!\!\!\perp e_{X3}\), and
\(e_{X2} \perp\!\!\!\perp e_{X3}\).

@@

\section{The probabilistic implications of DAGs}\label{sec-appendix-D}

After completing the identification process outlined in
Section~\ref{sec-appendix-B}, researchers estimate the ACE using
Bayesian inference methods. Bayesian inference calculates the
probability of a set of parameters \(\theta\), enabling researchers to
use their distributions to represent the ACE. The approach begins by
defining two probability distributions: the likelihood of the data,
\(P(X_{1}, X_{2}, \dots, X_{n} | \theta)\), and the prior distribution,
\(P(\theta)\) \citep{Everitt_et_al_2010}. Here, \(X_{n}\) represents a
random variable, and for simplicity, \(\theta\) denotes a parameter
space of dimension one. After observing empirical data, researchers
update these priors to posterior distributions using Bayes' rule
\citep{Jeffreys_1998}:

\begin{equation}\phantomsection\label{eq-bayes_rule}{
P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
}\end{equation}

Researchers can further simplify the posterior updating process in
Equation~\ref{eq-bayes_rule} into two steps: integrating new empirical
data, which is defined by the likelihood, with the parameter update,
which is determined by the priors, as demonstrated in equation
Equation~\ref{eq-prop_rule}. This simplification is possible because the
denominator on the right-hand side of Equation~\ref{eq-bayes_rule} acts
as a normalizing constant, independent of \(\theta\).

\begin{equation}\phantomsection\label{eq-prop_rule}{
P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
}\end{equation}

Temporarily setting aside the definition of prior distributions, it is
important to note that the posterior updating process depends on the
assumptions underlying the likelihood of the data. However, as the
number of random variables, \(n\), increases, this joint distribution
becomes quickly intractable \citep{Neal_2020}. This is evident from
Equation~\ref{eq-like_chain}, where the likelihood distribution is
re-expressed in terms of chained conditional distributions.

\begin{equation}\phantomsection\label{eq-like_chain}{
P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
}\end{equation}

To address the complexity of the likelihood, researchers simplify the
distribution by assuming that variables exhibit specific local
(in)dependencies. These assumptions improve model tractability and
streamline the estimation process. Researchers formalize these local
(in)dependencies using directed acyclic graphs (DAGs). DAGs represent
the probabilistic implications of these assumptions by relying on three
key principles: the local Markov assumption, the minimality assumption,
and the causal edges assumption. The \emph{local Markov assumption}
defines the statistical independencies implied by a DAG. It states that
a variable is independent of all its non-descendants, given its parents
\citep{Neal_2020}. The \emph{minimality assumption} identifies the
statistical dependencies implied by a DAG. It asserts that every pair of
adjacent variables are dependent \citep{Neal_2020}. Finally, the
\emph{causal edges assumption} establishes causal relationships between
parents and their children in a DAG by stating that every parent is a
direct cause of their children \citep{Neal_2020}. Figure
Figure~\ref{fig-ACflow} illustrates the flow of association and
causation in graphs.

@@

This lead to the \emph{Bayesian Network factorization} of
Equation~\ref{eq-like_chain}.

\begin{equation}\phantomsection\label{eq-net_factor}{
P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
}\end{equation}

>>>>

In this section we focus on the identification of several parameters of
interest from the theoretical model, together with various assessment
design features relevant to CJ experiments. The main reason to focus on
identification is that if these parameters cannot be identified using
``infinite'' and error free data, we certainly cannot learn anything
about it using finite data. So, for the time being, we assume large
samples and error-free measurements \citep[pp.~5]{Schuessler_et_al_2023}

@@

\includegraphics[width=0.23\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_01.png}

\caption{\label{fig-CJ_TM_01}}

\includegraphics[width=0.32\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_02.png}

\caption{\label{fig-CJ_TM_02}}

\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}

\caption{\label{fig-CJ_TM_03}}

@@

\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_05.png}

\caption{\label{fig-CJ_TM_05}}

\includegraphics[width=0.59\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_06.png}

\caption{\label{fig-CJ_TM_06}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_07.png}

\caption{\label{fig-CJ_TM_07}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_08.png}

\caption{\label{fig-CJ_TM_08}}

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}

\caption{\label{fig-CJ_TM_09}}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}

\caption{\label{fig-CJ_TM_11}}

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}

\caption{\label{fig-CJ_TM_13}}

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}

\caption{\label{fig-CJ_TM_15}}

@@

Researchers would then proceed to identify the estimands.
\emph{Identification} is concerned with determining whether an estimator
can accurately compute an estimand based on its assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} is a method or function that transforms data into an
estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand and are derived through \emph{estimation},
which refers to the process of integrating data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
visual representation of the process of transitioning from estimands to
estimates.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Assuming the depicted causal structure is correct, SCMs and DAGs serve
as \emph{conceptual (theoretical) models} that help researchers
determine which statistical models can identify an estimand, thus
enabling valid causal inference \citep{McElreath_2020}. They achieve
this by representing various causal structures using only five
fundamental building blocks, regardless of complexity
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components
\citep{McElreath_2020}. Furthermore, SCMs and DAGs can depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
requiring detailed specifications of the functional relating the
variables, their parameters, or their data types
\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows how theoretical
models guide the inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs indicate
the expected direction of causal influences between these variables.
While DAGs often omit exogenous variables for simplicity (the
\emph{standard} representation), including these variables is
advantageous (the \emph{magnified} representation shown in the figures),
as their inclusion helps to identify potential issues related to
conditioning and confounding \citep{Cinelli_et_al_2020}.

@@

A careful examination of the figures highlights the theoretical
assumptions underlying these building blocks. Figures \ref{fig-scm_bb1}
and \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
scenario where variables \(X_{1}\) and \(X_{3}\) are not causally
related. Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
connected nodes, representing a scenario where a \emph{parent} node
\(X_{1}\) exerts a causal influence on a \emph{child} node \(X_{3}\). In
this setup, \(X_{3}\) is considered a \emph{descendant} of \(X_{1}\).
Additionally, \(X_{1}\) and \(X_{3}\) are described as \emph{adjacent}
because an edge directly connects them. Figures \ref{fig-scm_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X_{1}\) influences
\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In this configuration,
\(X_{1}\) is a parent node of \(X_{2}\), which is a parent node of
\(X_{3}\). This structure creates a \emph{directed path} between
\(X_{1}\) and \(X_{3}\). Consequently, \(X_{1}\) is an \emph{ancestor}
of \(X_{3}\), and \(X_{2}\) fully \emph{mediates} the relationship
between the two. Figures \ref{fig-scm_bb4} and \ref{fig-mdag_bb4}
illustrate a \emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are
both influenced by \(X_{2}\). Here, \(X_{2}\) is a parent node that
\emph{confounds} the relationship between \(X_{1}\) and \(X_{3}\).
Finally, Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a
\emph{collider}, where variables \(X_{1}\) and \(X_{3}\) are concurrent
causes of \(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are
not causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, in all SCMs, the errors are assumed to be
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \perp\!\!\!\perp e_{X2}\),
\(e_{X1} \perp\!\!\!\perp e_{X3}\), and
\(e_{X2} \perp\!\!\!\perp e_{X3}\).

@@

\section{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-D}

In addition to their graphical capabilities, SCMs and DAGs can encode
probabilistic information about causal structures. They achieve this
encoding by relying on three fundamental assumptions: the local Markov,
the minimality, and the causal edges assumption. The \emph{local Markov
assumption} encodes probabilistic independencies between variables by
declaring that nodes in a graph are independent of all its
non-descendants, given its parents \citep{Neal_2020}. Meanwhile, the
\emph{minimality assumption} encodes probabilistic dependencies among
variables by stating that every pair of adjacent nodes exhibits a
dependency \citep{Neal_2020}. Finally, the \emph{causal edges
assumption} encodes causal relationships between variables by declaring
that each parent node acts as a direct cause of its children
\citep{Neal_2020}. Figure~\ref{fig-ACflow} illustrates how these
assumptions influence the statistical and causal interpretations of
graphs.

@@

The encoding of probabilistic information in SCMs and DAGs facilitates
the efficient representation of the joint distribution of variables
\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
assumptions underlying this encoding is that any conceptual model
described by an SCM and DAG can express the joint distribution of
variables as a product of conditional probability distributions (CPDs)
in the form \(P(child \: | \: parents)\). This property is formally
known as the \emph{Bayesian Network factorization}
(Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this factorization, \(pa(X_{i})\) denotes the set of variables that
are the parents of \(X_{i}\).

\begin{equation}\phantomsection\label{eq-net_factor}{
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
}\end{equation}

Ultimately, this encoding enables researchers with conceptual
(theoretical) knowledge in the form of an SCM and DAG to predict
patterns of (in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model.

The concept of predicting patterns of (in)dependencies in the data is
more clearly illustrated through the five fundamental building blocks
outlined in Section~\ref{sec-appendix-C}.
